---
title: "Replication of Keyword, STM, and Empath Figures"
format: html
execute:
  echo: false
  warning: false
  message: false
---

This document reproduces keyword, STM, and Empath analyses from the replication pipeline.

```{r}
library(targets)
library(tidyverse)
library(stm)
library(ggridges)

tar_load(hot100_genius_full)
```

# Overview

This document reproduces the keyword, STM, and Empath figures and tables referenced in the manuscript using the current pipeline data. It mirrors the structure of the paper, with Quarto labels suitable for citation in text.

Note: Figure 1 in the paper is a screenshot of a Billboard chart page and is not programmatically reproduced here.

# Data Preparation

```{r}
lyrics_all <- hot100_genius_full |>
  mutate(
    lyrics_clean = stringr::str_to_lower(dplyr::coalesce(lyrics, "")),
    lyrics_clean = stringr::str_replace_all(lyrics_clean, "[^[:alnum:]]", " ")
  )

lyrics_matched <- lyrics_all |>
  filter(!is.na(lyrics), lyrics != "", !is.na(match_score), match_score < 0.3)
```

# Figure 2: Missing Lyrics by Year (Paper Figure 2)

```{r}
#| label: fig-missing-lyrics
#| fig-cap: "Missing song lyrics in Genius, by year (supplementary material)."

missing_by_year <- lyrics_all |>
  group_by(year) |>
  summarise(missing = mean(is.na(lyrics) | lyrics == ""), .groups = "drop")

ggplot(missing_by_year, aes(x = year, y = missing)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Year", y = "Proportion missing")
```

# Figure 3: Political Keyword Search Results (Paper Figure 3)

```{r}
#| label: fig-keyword-politics
#| fig-cap: "Political keyword search results (manuscript Figure 1)."

political_terms <- tibble::tribble(
  ~label, ~pattern,
  "Politics / Political", "\\bpolitic(s|al)\\b",
  "War", "\\bwar(s)?\\b",
  "Democra*", "democra",
  "Vote / Voting", "\\bvote\\b|\\bvoting\\b"
)

political_series <- political_terms |>
  mutate(
    series = map(pattern, \(pat) {
      lyrics_matched |>
        group_by(year) |>
        summarise(prop = mean(str_detect(lyrics_clean, pat)), .groups = "drop")
    })
  ) |>
  unnest(series)

ggplot(political_series, aes(x = year, y = prop)) +
  geom_line() +
  facet_wrap(~label, ncol = 2, scales = "free_y") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Year", y = "Proportion of songs") +
  theme_bw()
```

# Figure 4: Country Keyword Search Results (Paper Figure 4)

```{r}
#| label: fig-keyword-war-countries
#| fig-cap: "Country keyword search results (supplementary material)."

war_terms <- tibble::tribble(
  ~label, ~pattern,
  "Korea", "\\bkorea\\b",
  "Vietnam", "\\bvietnam\\b",
  "Iraq", "\\biraq\\b",
  "Afghanistan", "\\bafghan\\w*\\b"
)

war_series <- war_terms |>
  mutate(
    series = map(pattern, \(pat) {
      lyrics_matched |>
        group_by(year) |>
        summarise(prop = mean(str_detect(lyrics_clean, pat)), .groups = "drop")
    })
  ) |>
  unnest(series)

ggplot(war_series, aes(x = year, y = prop)) +
  geom_line() +
  facet_wrap(~label, ncol = 2) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.01)) +
  labs(x = "Year", y = "Proportion of songs") +
  theme_bw()
```

# Table 1: Highest Probability Words (Paper Table 1)

```{r}
#| label: tbl-stm-topics
#| tbl-cap: "Highest probability words from the STM topics (manuscript Table 2)."

if ("stm_topic_labels" %in% targets::tar_meta()$name) {
  tar_load(stm_topic_labels)

  stm_prob <- stm_topic_labels$prob
  colnames(stm_prob) <- paste0("Word ", seq_len(ncol(stm_prob)))

  stm_topic_table <- as_tibble(stm_prob) |>
    mutate(Topic = row_number()) |>
    relocate(Topic)

  knitr::kable(stm_topic_table)
} else {
  knitr::kable(
    tibble::tibble(Note = "STM topic labels not available. Run STM pipeline to reproduce Table 1.")
  )
}
```

# Figure 5: STM Conflict and Violence Over Time (Paper Figure 5)

```{r}
#| label: fig-stm-violence
#| fig-cap: "STM conflict and violence topics over time (supplementary material)."

if (all(c("stm_effects", "stm_topic_labels") %in% targets::tar_meta()$name)) {
  tar_load(stm_effects)
  tar_load(stm_topic_labels)

  violence_terms <- c("die", "fight", "kill", "dead", "gun", "blood", "war", "violence")
  topic_words <- stm_topic_labels$prob
  topic_scores <- apply(
    topic_words,
    1,
    \(words) sum(tolower(words) %in% violence_terms)
  )
  violence_topic <- which.max(topic_scores)
  if (length(violence_topic) == 0 || max(topic_scores) == 0) {
    violence_topic <- 70
  }

  if ("stm_fit" %in% targets::tar_meta()$name) {
    tar_load(stm_fit)
    plot(
      stm_effects,
      "Year",
      method = "continuous",
      linecol = "black",
      printlegend = FALSE,
      xlab = "Year",
      main = paste0("Conflict and Violence Topic (Topic ", violence_topic, ")"),
      topics = violence_topic,
      model = stm_fit
    )
  } else {
    plot(
      stm_effects,
      "Year",
      method = "continuous",
      linecol = "black",
      printlegend = FALSE,
      xlab = "Year",
      main = paste0("Conflict and Violence Topic (Topic ", violence_topic, ")"),
      topics = violence_topic
    )
  }
} else {
  ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = "STM results not available. Run the STM pipeline to reproduce this figure."
    ) +
    theme_void()
}
```

# Figure 6: Empath Results Over Time (Paper Figure 6)

```{r}
#| label: fig-empath-over-time
#| fig-cap: "Empath results over time (supplementary material)."

if ("empath_results" %in% targets::tar_meta()$name) {
  tar_load(empath_results)

  empath_long <- bind_rows(
    empath_results |> filter(politics > 0) |> count(year) |> mutate(category = "Politics"),
    empath_results |> filter(politics_nytimes > 0) |> count(year) |> mutate(category = "Politics (NYTimes)"),
    empath_results |> filter(politics_reddit > 0) |> count(year) |> mutate(category = "Politics (Reddit)"),
    empath_results |> filter(politics_fiction > 0) |> count(year) |> mutate(category = "Politics (Fiction)")
  )

ggplot(empath_long, aes(x = year, y = n)) +
  geom_line() +
  facet_wrap(~category, ncol = 2, scales = "free_y") +
  labs(x = "Year", y = "Songs with non-zero score")
} else if (file.exists(here::here("empath", "empath_politics_freq.pdf"))) {
  knitr::include_graphics(here::here("empath", "empath_politics_freq.pdf"))
} else {
  ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = "Empath results not available. Run the Empath pipeline to reproduce this figure."
    ) +
    theme_void()
}
```

# Table 2: Empath Results, Top Songs (Paper Table 2)

```{r}
#| label: tbl-empath-top-songs
#| tbl-cap: "Empath results for top songs (manuscript Table 3)."

if ("empath_results" %in% targets::tar_meta()$name) {
  tar_load(empath_results)

  empath_unique <- empath_results |>
    select(
      artist,
      title,
      politics,
      politics_nytimes,
      politics_reddit,
      politics_fiction,
      government,
      independence,
      violence,
      military,
      terrorism,
      law
    ) |>
    group_by(artist, title) |>
    summarise(across(where(is.numeric), ~ max(.x, na.rm = TRUE)), .groups = "drop")

  top_empath <- function(column, panel) {
    empath_unique |>
      filter(.data[[column]] > 0) |>
      arrange(desc(.data[[column]])) |>
      slice_head(n = 6) |>
      transmute(
        Panel = panel,
        Artist = artist,
        Title = title,
        Score = round(.data[[column]], 3)
      )
  }

  empath_table_2 <- bind_rows(
    top_empath("politics", "Panel A: Politics"),
    top_empath("politics_nytimes", "Panel B: Politics (NY Times)"),
    top_empath("politics_reddit", "Panel C: Politics (Reddit)"),
    top_empath("politics_fiction", "Panel D: Politics (Fiction)")
  )

  knitr::kable(empath_table_2)
} else {
  knitr::kable(
    tibble::tibble(Note = "Empath results not available. Run Empath pipeline to reproduce Table 2.")
  )
}
```

# Table 3: Empath Results, New Politics Measures (Paper Table 3)

```{r}
#| label: tbl-empath-top-songs-expanded
#| tbl-cap: "Empath results for expanded politics measures (supplementary material)."

if ("empath_results" %in% targets::tar_meta()$name) {
  tar_load(empath_results)

  empath_unique <- empath_results |>
    select(
      artist,
      title,
      politics,
      politics_nytimes,
      politics_reddit,
      politics_fiction,
      government,
      independence,
      violence,
      military,
      terrorism,
      law
    ) |>
    group_by(artist, title) |>
    summarise(across(where(is.numeric), ~ max(.x, na.rm = TRUE)), .groups = "drop")

  top_empath <- function(column, panel) {
    empath_unique |>
      filter(.data[[column]] > 0) |>
      arrange(desc(.data[[column]])) |>
      slice_head(n = 6) |>
      transmute(
        Panel = panel,
        Artist = artist,
        Title = title,
        Score = round(.data[[column]], 3)
      )
  }

  empath_table_3 <- bind_rows(
    top_empath("government", "Panel A: Government"),
    top_empath("independence", "Panel B: Independence"),
    top_empath("violence", "Panel C: Violence"),
    top_empath("military", "Panel D: Military"),
    top_empath("terrorism", "Panel E: Terrorism"),
    top_empath("law", "Panel F: Law")
  )

  knitr::kable(empath_table_3)
} else {
  knitr::kable(
    tibble::tibble(Note = "Empath results not available. Run Empath pipeline to reproduce Table 3.")
  )
}
```

# Table 4: Empath Scores for Known Political Songs (Paper Table 4)

```{r}
#| label: tbl-empath-known-songs
#| tbl-cap: "Empath scores for known political songs (manuscript Table 4)."

if ("empath_results" %in% targets::tar_meta()$name) {
  tar_load(empath_results)

  known_politics <- readr::read_csv(
    "data-raw/known_politics_songs.csv",
    show_col_types = FALSE
  )

  empath_score_cols <- c(
    "politics",
    "politics_nytimes",
    "politics_reddit",
    "politics_fiction",
    "government",
    "independence",
    "violence",
    "military",
    "terrorism",
    "law"
  )

  normalize_join_key <- function(x) {
    x |>
      stringr::str_to_lower() |>
      stringr::str_replace_all("[^[:alnum:]]+", " ") |>
      stringr::str_squish()
  }

  summarise_empath_scores <- function(data) {
    data |>
      select(any_of(c("artist", "title", empath_score_cols))) |>
      group_by(artist, title) |>
      summarise(
        across(
          all_of(empath_score_cols),
          \(x) if (all(is.na(x))) NA_real_ else max(x, na.rm = TRUE)
        ),
        .groups = "drop"
      )
  }

  empath_scored_filtered <- summarise_empath_scores(empath_results) |>
    mutate(
      artist_key = normalize_join_key(artist),
      title_key = normalize_join_key(title),
      source_priority = 1L
    )

  empath_scored_raw <- readr::read_csv(
    "empath/output/empath_using_seeds.csv",
    show_col_types = FALSE
  ) |>
    select(-1) |>
    summarise_empath_scores() |>
    mutate(
      artist_key = normalize_join_key(artist),
      title_key = normalize_join_key(title),
      source_priority = 2L
    )

  empath_scored <- bind_rows(empath_scored_filtered, empath_scored_raw) |>
    arrange(artist_key, title_key, source_priority) |>
    group_by(artist_key, title_key) |>
    slice_head(n = 1) |>
    ungroup() |>
    select(-source_priority, -artist, -title)

  known_scores <- known_politics |>
    mutate(
      artist_key = normalize_join_key(artist),
      title_key = normalize_join_key(title)
    ) |>
    left_join(empath_scored, by = c("artist_key", "title_key")) |>
    mutate(across(all_of(empath_score_cols), ~ tidyr::replace_na(.x, 0))) |>
    arrange(year, artist, title) |>
    select(
      Artist = artist,
      Title = title,
      Year = year,
      Politics = politics,
      `Politics (NY Times)` = politics_nytimes,
      `Politics (Reddit)` = politics_reddit,
      `Politics (Fiction)` = politics_fiction,
      Government = government,
      Independence = independence,
      Violence = violence,
      Military = military,
      Terrorism = terrorism,
      Law = law
    ) |>
    mutate(across(where(is.numeric), ~ round(.x, 3)))

  knitr::kable(known_scores)
} else {
  knitr::kable(
    tibble::tibble(Note = "Empath results not available. Run Empath pipeline to reproduce Table 4.")
  )
}
```
